---
title: "Learning ggplot2"
author: "Lina Andersson"
date: "2026-09-18 11:37"
date-format: "YYYY-MM-DD HH:mm"
categories: [Instructions, ggplot2]
---

## Learning ggplot2
Here, you will learn how to make plots using the R package ggplot2. To learn this we will follow the tutorial by [Raukr](https://nbisweden.github.io/raukr-2026/labs/ggplot/):

### 1. Basics
1. First, we need to load our required packages:
```{r}
#| eval: false
#| warning: false
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggrepel)
```

2. For this example, we will use the `iris dataset`:
```{r}
#| warning: false
data("iris")
head(iris)
```

3. Next, we build a plot with x= `Petal length` and y= `Petal Width`:

When you want to specify which part of your data to use you use `Aesthetic mapping`. And we specify the geometric using geoms. Note that the variable names do not have double quotes "" like in base plots.


```{r}
#| warning: false
library(ggplot2) #OBS! In Quarto you need to load ggplot2 in the first code chunk using ggplot2
ggplot(data = iris) +
  geom_point(mapping = aes(x = Petal.Length, y = Petal.Width))
```

4. 