---
title: "Learning ggplot2"
author: "Lina Andersson"
date: "2026-09-18 11:37"
date-modified: "2026-09-24 14:05"
date-format: "YYYY-MM-DD"
time: "11:37"
categories: [Instructions, ggplot2]
---

## Learning ggplot2
Here, you will learn how to make plots using the R package ggplot2. To learn this we will follow the tutorial by [Raukr](https://nbisweden.github.io/raukr-2026/labs/ggplot/):

There are two exercises apart from learning basics, in which we will cover one in this blog post:

### Economist scatterplot

The aim of this challenge is to recreate the plot below originally published in [*The Economist*](https://www.economist.com/graphic-detail/2011/12/02/corrosive-corruption). The graph is a scatterplot showing the relationship between *Corruption Index and Human Development Index for various countries*.

```{r}
#| warning: false
#| echo: false
# load packages
library(ggrepel)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(showtext)
showtext_auto()
font_add_google("Slabo 27px", "Slabo 27px")

# read data
ec <- read.csv("data_economist.csv", header = TRUE)
# refactor
ec$Region <- factor(ec$Region,
  levels = c(
    "EU W. Europe", "Americas", "Asia Pacific",
    "East EU Cemt Asia", "MENA", "SSA"
  ),
  labels = c(
    "OECD", "Americas", "Asia &\nOceania",
    "Central &\nEastern Europe",
    "Middle East &\nNorth Africa",
    "Sub-Saharan\nAfrica"
  )
)

# labels
labels <- c("Congo", "Afghanistan", "Sudan", "Myanmar", "Iraq", "Venezuela", "Russia", "Argentina", "Brazil", "Italy", "South Africa", "Cape Verde", "Bhutan", "Botswana", "Britain", "New Zealand", "Greece", "China", "India", "Rwanda", "Spain", "France", "United States", "Japan", "Norway", "Singapore", "Barbados", "Germany")

# trendline label
trend_model <- lm(HDI ~ poly(CPI, 2), data = ec)
r_squared <- summary(trend_model)$r.squared
r_label <- as.expression(bquote(R^2 == .(round(r_squared, 2))))

# plotting
p1 <- ggplot(ec, aes(x = CPI, y = HDI, color = Region)) +
  geom_smooth(aes(fill = "red"), method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "red", linewidth = 0.6) +
  geom_point(shape = 21, size = 3, stroke = 0.8, fill = "white") +
  geom_text_repel(
    data = subset(ec, Country %in% labels), aes(label = Country),
    color = "black", box.padding = unit(1, "lines"), segment.size = 0.25,
    size = 3, family = "Slabo 27px"
  ) +
  scale_x_continuous(
    name = "Corruption Perceptions Index, 2011 (10=least corrupt)",
    breaks = 1:10, limits = c(1, 10)
  ) +
  scale_y_continuous(
    name = "Human Development Index, 2011 (1=best)",
    breaks = seq(from = 0.2, to = 1, by = 0.1), limits = c(0.2, 1)
  ) +
  scale_color_manual(values = c("#23576E", "#099FDB", "#29B00E", "#208F84", "#F55840", "#924F3E")) +
  scale_fill_manual(name = "trend", values = "red", labels = r_label) +
  labs(
    title = "Corruption and human development",
    caption = "Sources: Transparency International; UN Human Development Report"
  ) +
  guides(color = guide_legend(nrow = 1)) +
  theme_bw(base_family = "slabo") +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    legend.title = element_blank(),
    axis.title = element_text(face = "italic"),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_line(color = "grey60"),
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(hjust = 0, size = 8)
  )
  p1
```

**1. First, we need to load our required packages:**
```{r}
#| eval: false
#| warning: false
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggrepel)
```

**2. Download the [data .csv file](../ggplot/data_economist.csv) and read the data using:**
```{r}
#| warning: false
ec <- read.csv("data_economist.csv", header = TRUE)
head(ec)
```

To check that the right types of data is in the correct format we use:

* x = 'CPI' (numeric)
* y = 'HDI' (numeric)
* 'Region' (factor)

```{r}
#| warning: false
str(ec)
```

**3. 'Region' is currently formatted as `character`. We change this by running:**
```{r}
#| warning: false
ec$Region <- factor(ec$Region)
levels(ec$Region)
```

**4. Change names of 'Region':**

The names of the regions also need to be changed, since the new labels are a bit long for one line we can add `\n` where we want breaks:
```{r}
#| warning: false
ec$Region <- factor(ec$Region,
  levels = c(
    "EU W. Europe",
    "Americas",
    "Asia Pacific",
    "East EU Cemt Asia",
    "MENA",
    "SSA"
  ),
  labels = c(
    "OECD",
    "Americas",
    "Asia &\nOceania",
    "Central &\nEastern Europe",
    "Middle East &\nNorth Africa",
    "Sub-Saharan\nAfrica"
  )
)

levels(ec$Region)
```

**5. Start building the plot using ggplot2:**

The color of the points should be on 'Region'. 
```{r}
ggplot(ec, aes(x = CPI, y = HDI, color = Region)) +
  geom_point(shape = 21, size = 3, stroke = 0.8, fill = "white")
```

**6. Add a trendline:** 

The line should be red, follow a custom equation `y ~ poly(x, 2)` for the linear model (lm), and not include confidence interval shading (se = Standard Error).
```{r}
ggplot(ec, aes(x = CPI, y = HDI, color = Region)) +
  geom_point(shape = 21, size = 3, stroke = 0.8, fill = "white") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, linewidth = 0.6, color = "red")
```

**7. Add a legend for trendline and layer it underneath points:**

We need to reorder the geoms and create a fake legend by adding aes(). 
```{r}
p <- ggplot(ec, aes(x = CPI, y = HDI, color = Region)) +
  geom_smooth(aes(fill = "red"), method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "red", linewidth = 0.6) +
  geom_point(shape = 21, size = 3, stroke = 0.8, fill = "white")
```

**8. Add text labels for points:**

To add labels for the points we need to make a list of names for the points by using concatenation of the names. In addition, since we gave the previous plot the variable `p` we can simply add the labels after using `geom_text()`.
```{r}
labels <- c("Congo", "Afghanistan", "Sudan", "Myanmar", "Iraq", "Venezuela", "Russia", "Argentina", "Brazil", "Italy", "South Africa", "Cape Verde", "Bhutan", "Botswana", "Britain", "New Zealand", "Greece", "China", "India", "Rwanda", "Spain", "France", "United States", "Japan", "Norway", "Singapore", "Barbados", "Germany")

p + geom_text(data = subset(ec, Country %in% labels), aes(label = Country), color = "black")
```

**9. Change to a custom font:**

Custom fonts can be used for the labels by providing the font name to argument `family` like so `geom_text(family="fontname")`. This is where the `showtext` package comes into play. The package registers fonts from 'Google Fonts' or a local font file. 

The font used in *The Economist* is not free, we use a font called `Slabo 27px` instead.
```{r}
p + geom_text(
  data = subset(ec, Country %in% labels), aes(label = Country),
  color = "black", family = "Slabo 27px"
)
```

**10. Correcting label overlap:**

To avoid overlapping of labels, we can use a ggplot2 extension package `ggrepel`. We can use function `geom_text_repel()`.

```{r}
p <- p + geom_text_repel(
  data = subset(ec, Country %in% labels), aes(label = Country),
  color = "black", box.padding = unit(1, "lines"), segment.size = 0.25,
  size = 3, family = "Slabo 27px"
)
p
```

**11. Adjust axes:**

We can adjust axes breaks, axes labels, point colors and relabeling the trendline legend text.

* x ='Corruption Perceptions Index, 2011 (10=least corrupt)'
* y = 'Human Development Index, 2011 (1=best)'

* x axis breaks = 1 to 10 by 1 increment
* y axis breaks = 0.2 to 1.0 by 0.1 increments

```{r}
p <- p + scale_x_continuous(
  name = "Corruption Perceptions Index, 2011 (10=least corrupt)",
  breaks = 1:10, limits = c(1, 10)
) +
  scale_y_continuous(
    name = "Human Development Index, 2011 (1=best)",
    breaks = seq(from = 0.2, to = 1, by = 0.1), limits = c(0.2, 1)
  )
p
```

**12. Pick colors manually:**

For this we use two functions:

* `scale_color_manual()` to set the color codes for points
* `scale_fill_manual()` to change the trendline label. Here we compute the model fit first, then use the resulting value to create the legemd entry as `R²`

```{r}
trend_model <- lm(HDI ~ poly(CPI, 2), data = ec)
r_squared <- summary(trend_model)$r.squared
r_label <- as.expression(bquote(R^2 == .(round(r_squared, 2))))

p <- p + scale_color_manual(values = c("#23576E", "#099FDB", "#29B00E", "#208F84", "#F55840", "#924F3E")) +
  scale_fill_manual(name = "trend", values = "red", labels = r_label)
p
```

**13. Add a Title and Captions:**

Set these using the function `labs()`.

* Title = 'Corruption and human development'
* Caption = 'Sources: Transparency International; UN Human Development Report'

```{r}
p <- p + labs(
  title = "Corruption and human development",
  caption = "Sources: Transparency International; UN Human Development Report"
)
p
```

**14. Add a Theme:**

The theme we want is the following:

* Move the legend to the top and as a single row using `legend_position`
* Set a custom font for all text elements using using `base_family`


```{r}
p <- p + guides(color = guide_legend(nrow = 1)) +
  theme_bw(base_family = "Slabo 27px") +
  theme(legend.position = "top")
p
```

**15. Final changes:**

* Turn off minor gridlines
* Turn off major gridlines on x-axis
* Remove the gray background
* Remove panel border
* Remove legend titles
* Make axes titles italic
* Turn off y-axis ticks
* Change x-axis ticks to color grey60
* Make plot title bold
* Decrease size of caption to size 8


```{r}
p + theme(
  panel.grid.minor = element_blank(),
  panel.grid.major.x = element_blank(),
  panel.background = element_blank(),
  panel.border = element_blank(),
  legend.title = element_blank(),
  axis.title = element_text(face = "italic"),
  axis.ticks.y = element_blank(),
  axis.ticks.x = element_line(color = "grey60"),
  plot.title = element_text(face = "bold"),
  plot.caption = element_text(hjust = 0, size = 8)
)
```
